Symbolic Abstract Data Type Inference

### 2015-07-07 appendix.tex
Section titles + restructuring the restructuring.

The result follows from the fact that $\preceq$ is a well-quasi-ordering on width-bounded histories.
It is well-known that every potentially infinite set of elements has a finite set of minimals w.r.t. a well-quasi-ordering.

A \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
elements from $X$, there exists indices $i < j$ such that $R(x_i,x_j)$.

\begin{example}

  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
  {\tt pop} operation with undefined matching. Because each successive $h_i$
  has both more matches and more pending operations, no two histories of the
  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.

\end{example}

The previous example demonstrates that $\preceq$ is not a wqo by allowing each
history $h_i$ of the infinite sequence to contain more and more pending
operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
Curbing this ability by limiting the maximum amount of pending operations per
history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
essentially limits us to width-$k$ histories, of executions with at most $k$
operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
this restriction comes at no loss of completeness when considering only the
histories of bounded-width ADT kernels, which is the subject of the remainder
of this section.

>>>>

It is well-known that every potentially infinite set of elements has a finite set of minimals w.r.t. a well-quasi-ordering.

### 2015-07-07 formula.tex
Section titles + restructuring the restructuring.

%!TEX root = draft.tex
\section{Computing Symbolic Representations}
\label{sec:formula}

While Section~\ref{sec:patterns} demonstrates that (the kernel complements) of
naturally-occurring ADTs can be represented finitely, in this section we
demonstrate that such finite representations have logical interpretations,
allowing us to derive the formulas representing ADTs. In what follows, we
describe how to obtain a formula which is satisfied by the histories which
embed a given history, via $\preceq$. Then, using the finite set of histories
which represent the complement of a given ADT~\footnote{The complement of the kernel of an ADT $A$ is simply called the complement of $A$.}, we represent the ADT itself as
the conjunction of negations of these embedding formulas. The resulting formula
is satisfied only on the histories which do not embed the generators of a given
ADT’s complement.

>>>>

%!TEX root = draft.tex
\section{Symbolic ADT Representations}
\label{sec:formula}

While Section~\ref{sec:patterns} demonstrates that (the kernel complements) of
naturally-occurring ADTs can be represented finitely, in this section we
demonstrate that such finite representations have logical interpretations,
allowing us to derive the formulas representing ADTs. In what follows, we
describe how to obtain a formula which is satisfied by the histories which
embed a given history, via $\preceq$. Then, using the finite set of histories
which represent the complement of a given ADT, we represent the ADT itself as
the conjunction of negations of these embedding formulas. The resulting formula
is satisfied only on the histories which do not embed the generators of a given
ADT’s complement.

### 2015-07-07 patterns.tex
Section titles + restructuring the restructuring.

%!TEX root = draft.tex
\section{Computing Finite ADT Representations}
\label{sec:patterns}

In this section we demonstrate that naturally-occurring ADTs can be precisely
represented by finite sets of histories, despite the fact that these ADTs admit
infinite sets of histories. This results is nontrivial, and relies on the
identification of certain algebraic properties of the sets of histories
admitted by ADTs, more precisely, that they are
closed under the removal of certain operations. 
%and that the bounded complement of their kernel has 
%sets of histories
%%adhere to a well-founded ordering under the relation induced by such removals.
%have 
%a finite set of minimal elements with respect to the relation induced by such removals.

In addition to the relations $\to_\mathrm{o}$, $\to_\mathrm{p}$, and
$\to_\mathrm{c}$ of Section~\ref{sec:histories} under which all ADTs are
closed, the ADTs we consider in this work are also closed under additional
relations which remove read-only operations, unmatched operations, entire
matches, and duplicate operations. The relations $\to_\mathrm{r}$,
$\to_\mathrm{u}$, $\to_\mathrm{m}$ and $\to_\mathrm{d}$ relate two histories
$h_1$ and $h_2$ when $h_2$ is obtained from $h_1$ by
\begin{itemize}

@@

\end{itemize}
We say an ADT closed under $\to_\mathrm{r}$, $\to_\mathrm{u}$, $\to_\mathrm{m}$
and $\to_\mathrm{d}$ is \emph{normal}. In Section~\ref{sec:nature} we
demonstrate that naturally-occurring ADTs are normal. Defining the relation
$\succeq$ as the reflexive and transitive closure of the above relations,
\begin{align*}
  \mathord{\succeq} = (
    \to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup 
    \to_\mathrm{r} \cup \to_\mathrm{u} \cup \to_\mathrm{m} \cup \to_\mathrm{d}
  )^\ast
\end{align*}
closure under $\succeq$ follows immediately.

\begin{lemma}\label{lem:ADT_closure}

@@

Besides this closure property, the $\preceq$ relation enjoys a certain notion
of well-foundedness (when restricted to a certain class of histories): every potentially 
infinite set of width-bounded histories has a finite set of minimal elements w.r.t. $\preceq$.

\begin{lemma}\label{lem:wqo}

  Sets of width-bounded histories have a finite set of $\preceq$-minimals.

@@

For the remainder of this section, we fix an ADT $A$, and let $B$ be the bounded
complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
$B$ has a finite set of $\preceq$-minimals. 
Furthermore, when $A$ normal, it is closed under
$\succeq$, and thus $B$ is closed under 
$\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$, and $h_2$ is width-bounded, then $h_2\in B$.}.
Closure under a relation satisfying Lemma~\ref{lem:wqo} implies
representation by a finite set. Formally, we say a set $X$ is \emph{finitely
representable} if there exists a finite set $Y$ and a relation $R \subseteq Y
\times X$ such that $X = \set{ x : \exists y\in Y.\ R(y,x) }$. In our case, we
obtain a finite set from which exactly the elements of $B$ are related by
$\preceq$.

>>>>

%!TEX root = draft.tex
\section{Finite ADT Representations}
\label{sec:patterns}

In this section we demonstrate that naturally-occurring ADTs can be precisely
represented by finite sets of histories, despite the fact that these ADTs admit
infinite sets of histories. This results relies on the identification of certain
algebraic properties of the sets of histories admitted by ADTs. In particular,
we find that sets of histories admitted by ADTs are closed under the removal
of certain operations, and that these sets adhere to a well-founded ordering
under the relation induced by such removals.

In addition to the relations $\to_\mathrm{o}$, $\to_\mathrm{p}$, and
$\to_\mathrm{c}$ of Section~\ref{sec:histories} under which all implementation
history sets are closed, the histories of ADT implementations we consider in
this work are also closed under additional relations which remove read-only
operations, unmatched operations, entire matches, and duplicate operations. The
relations $\to_\mathrm{r}$, $\to_\mathrm{u}$, $\to_\mathrm{m}$ and
$\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when $h_2$ is obtained
from $h_1$ by
\begin{itemize}

@@

\end{itemize}
We say an ADT whose histories are closed under $\to_\mathrm{r}$,
$\to_\mathrm{u}$, $\to_\mathrm{m}$ and $\to_\mathrm{d}$ is \emph{normal}. In
Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
normal. Defining the relation $\succeq$ as the reflexive and transitive closure
of the above relations,
\begin{align*}
  \mathord{\succeq} = (
    \to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup
    \to_\mathrm{r} \cup \to_\mathrm{u} \cup \to_\mathrm{m} \cup \to_\mathrm{d}
  )^\ast
\end{align*}
closure under $\succeq$ follows immediately.

\begin{lemma}
  \label{lem:ADT_closure}

@@

Besides this closure property, the $\preceq$ relation enjoys a certain notion of
well-foundedness when restricted to bounded-width histories: the set of
$\preceq$-minimal elements of any (potentially infinite) history set is finite.
This property is what enables us to represent infinite sets of invalid ADT
histories with a finite set of minimal examples. This property is captured
formally with wqos: a \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a
reflexive, transitive binary relation on $X$ for which in every infinite
sequence $x_0 x_1 \ldots$ of elements from $X$, there exists indices $i < j$
such that $R(x_i,x_j)$.

\begin{example}

  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
  {\tt pop} operation with undefined matching. Because each successive $h_i$
  has both more matches and more pending operations, no two histories of the
  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.

The previous example demonstrates that $\preceq$ is not a wqo by allowing each
history $h_i$ of the infinite sequence to contain more and more pending
operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
Curbing this ability by limiting the maximum amount of pending operations per
history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
essentially limits us to width-$k$ histories, of executions with at most $k$
operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
this restriction comes at no loss of completeness when considering only the
histories of bounded-width ADTs, which is the subject of the remainder
of this section.

\begin{lemma}
  \label{lem:wqo}

  $\preceq$ is a wqo on width-bounded histories.

@@

For the remainder of this section, we fix an ADT $A$, and let $B$ be its bounded
complement. When $A$ has bounded width, so does $B$, and thus $\preceq$ is a wqo
on $B$. Furthermore, when $A$ normal, it is closed under $\succeq$, and thus $B$
is closed under $\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when
restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$,
and $h_2$ is width-bounded, then $h_2\in B$.}. Closure under a relation
satisfying Lemma~\ref{lem:wqo} implies representation by a finite set. Formally,
we say a set $X$ is \emph{finitely representable} if there exists a finite set
$Y$ and a relation $R \subseteq Y \times X$ such that $X = \set{ x : \exists
y\in Y.\ R(y,x) }$. In our case, we obtain a finite set from which exactly the
elements of $B$ are related by $\preceq$.
